Liveness-Based Garbage Collection for Lazy Languages

### 2015-11-29 paper/paper.tex
latest - 29th

%% We consider a round of  garbage collection followed by a transition
%% as  a  step.  We  show  by  induction  that, for  programs  without
%% function  calls, %%  starting from  the %%  initial state  that has
%% $e_{\main}$ as the %% expression to be evaluated, any transition of
%% $n$ steps occurs without a \bang.

We first need an auxiliary  result about minefield semantics. Consider
a trace of  a minefield execution.  For every  minefield state $(\rho,
S,   \heap,  e(s),   \sigma)$   that   appears  on   the   LHS  of   a
$\rightsquigarrow$ step, the demand $\sigma$ on the expression $e$ (or
application $s$) is  non-null.  This can be proved by  an induction on
the number  of steps leading  to the  minefield state.  The  base step
holds   because  the   demand   $\sigma_{all}$   on  $(\mainpgm)$   is
non-null. For the inductive step we  observe that for each step of the
minefield semantics, if the demand $\sigma$  on the LHS of a minefield
step  is non-null,  the  demand  on the  RHS  is  a transformation  of
$\sigma$ (for example $\clazy\sigma$) which is also non-null.

%% Let  us clarify  this  observation: Since  the minefield  semantics
%% models lazy evaluation,  parts of the program may  not be evaluated
%% at all.  However,  if an expression (or application)  happens to be
%% evaluated, it will  appear on the LHS of a  minefield step, and the
%% context-sensitive demand on it will be non-null.

@@

\begin{lemma}
\label{lemma:call-less-cannot-go-bang}
Consider the minefield execution of  a program without function calls.
Such a program cannot enter the \bang\ state.
\end{lemma}
\begin{proof}
Call a $GC(\ldots)$  followed by a $\rightsquigarrow$  transition as a
step. Consider a state $(\rho, S,  \heap, e, \sigma)$ in the minefield
execution of a  program.  We show by induction on  the number of steps
$n$ leading  to this  state that  the next  transition cannot  enter a
\bang\ state.  When $n$ is 0,  the state is $([\;]_\rho,\_, [\;]_{S} ,
            [\;]_{\heap},  e_{\mainpgm},  \sigma_{all})$.   Since  the
            call to $GC(\ldots)$  in this state does  nothing, we just
            have to show that the $\rightsquigarrow$ transition cannot
            enter  a \bang\  state.  Since  our programs  are in  ANF,
            $e_{\mainpgm}$ can  only be  a $\LET$ expression.   A {\sc
              let}  step  does  not involve  dereferencing,  and  thus
            cannot result in a \bang.

@@

For the  {\sc car-1-clo}  rule, observe that  there are
two dereferences.  First $x$  is dereferenced to  get a
cons  cell  and then  the  head  of  the cons  cell  is
dereferenced  to  obtain  a  closure.   If  the  demand
$\sigma$ on  $(\CAR~x)$ is non-null, then  the liveness
of    $x$   will    include    both   $\epsilon$    and
$\acar\epsilon$,  and  a  garbage collection with  this  liveness  will
neither bind $x$ to a  $\bot$, nor insert $\bot$ at the
first   component  of   the   cons   cell.  Thus   both
dereferences    will    happen   normally    and    the
$\rightsquigarrow$ step will not enter \bang.
\end{proof}

>>>>

\subsection{Soundness Results}
We first need an auxiliary  result about minefield semantics. Consider
a trace of  a minefield execution.  For every  minefield state $(\rho,
S,   \heap,  e(s),   \sigma)$   that   appears  on   the   LHS  of   a
$\rightsquigarrow$ step, the demand $\sigma$ on the expression $e$ (or
application $s$) is  non-null.  This can be proved by  an induction on
the number  of steps leading  to the  minefield state.  The  base step
holds   because  the   demand   $\sigma_{all}$   on  $(\mainpgm)$   is
non-null. For the inductive step we  observe that for each step of the
minefield semantics, if the demand $\sigma$  on the LHS of a minefield
step  is non-null,  the  demand  on the  RHS  is  a transformation  of
$\sigma$ (for example $(\clazy\cup\acar)\sigma$) which is also non-null.

@@

\begin{lemma}
\label{lemma:call-less-cannot-go-bang}
Consider the minefield execution of  a program without function calls.
Such a program cannot enter the \bang\ state.
\end{lemma}
\begin{proof}
Call a $GC(\ldots)$  followed by a $\rightsquigarrow$  transition as a
step. Consider a state $(\rho, S,  \heap, e, \sigma)$ in the minefield
execution of a  program.  We show by induction on  the number of steps
$n$ leading  to this  state that  the next  transition cannot  enter a
\bang\ state.  When $n$ is 0,  the state is $([\;]_\rho,\,    (\rho_\mathit{init},   \, \ell_{\ans}  ,\,
(\print~\ans)):[\;]_{S}, 
[\;]_{\heap},\, (\mainpgm))$.   Since  the
            call to $GC$  in this state does  nothing, we just
            have to show that the $\rightsquigarrow$ transition cannot
            enter  a \bang\  state.  Since  our programs  are in  ANF,
            $e_{\mainpgm}$ can  only be  a $\LET$ expression.   A {\sc
              let}  step  does  not involve  dereferencing,  and  thus
            cannot result in a \bang.

@@

For the  {\sc car-1-clo}  rule, observe that  there are
two dereferences.  First $x$  is dereferenced to  get a
cons  cell  and then  the  head  of  the cons  cell  is
dereferenced  to  obtain  a  closure.   If  the  demand
$\sigma$ on  $(\CAR~x)$ is non-null, then  the liveness
of    $x$   will    include    both   $\epsilon$    and
$\acar\epsilon$,  and  a  garbage collection with  this  liveness  will
neither bind $x$ to a  $\bot$, nor insert $\bot$ at the
first   component  of   the   cons   cell.  Thus   both
dereferences    can take place without 
$\rightsquigarrow$ entering the   \bang\ state.
\end{proof}
